The binary liquid model for gravity-driven microchannel simulations

### 2010-10-08 Text/paper.tex
Added introduction about Bretherton problem and LBM description.

\section{Introduction}
Taylor/Bretherton \cite{bretherton} flow describes flow of long bubbles in
microchannels. It was found that the front meniscus film width is proportional
to $Ca^{2/3}$ for Capillary numbers less than $0.003$. While there are number
of works which simulate the Bretherton problem with the boundary value method
\cite{ingham-plates,heil-planar}, the simulation of Bretherton problem is of
difficulty for diffusion interface methods, where the interface is spread over
the few grid nodes. In this case the question of proper film width resolution
arises. While the discontinuous interface models are good to simulate
non-complicated geometries phenomena, the continuous interface methods are more
flexible to resolve problem evolving coalescence or droplet breakup. This work
concentrates on the resolving parameters range of the binary liquid lattice
Boltzmann method to validate it across the Bretherton problem.

@@

\section{Benchmark}

There are a number of the numerical methods which were used for the Taylor flow
simulation. \citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillar.
\citet{kreutzer-pressure-drop} performed numerical simulation by the
finite volume method the circular capillary for the number of different
Reynolds and capillary numbers. \citeauthor{wong-films} in
\cite{wong-films,wong-pressure} studied three-dimensional bubbles in the
polygonal capillaries and showed the derivations of different shapes in the
slug crossections and menisci appearance.
\citet{heil-bretherton,ingham-plates} studied the air finger propagation to
the two-dimensional channel for the range of Reynolds numbers and capillary
numbers. \citet{giavedoni-numerical} performed crossvalidation of the
finite element solution developed by them with the already published results
across the different capillary numbers. The solution is available for the
circular and planar cases.

While the applicability of other methods are well established for the
simulation of the Bretherton/Taylor problem, the lattice Boltzmann
applicability and thorough parameters study to the best authors' knowledge is
not yet done. One should acknowledge the works of
\citet{pagonabarraga-fingers} on meniscus in thin films for the fingering
phenomena. \citet{sehgal-microchannel} performed the lattice Boltzmann
simulations of two-dimensional channel flow for larger capillary numbers where
the authors found the discrepancies with the classical Bretherton theory having
limitations to the low capillary numbers regime \cite{giavedoni-numerical}.

\section{Lattice Botlzmann benchmark}
The purpose of this section is to establish the benchmark and perform
parameters study for the classical Bretherton/Taylor problem. There are certain
challenges for the lattice Boltzmann method which can be summarized as follows:
\begin{itemize}
 \item The diffusive interface nature of the multiphase model should not
influence the film resolution.
 \item The spurious currents arising from the surface tension discretization
can influence the mass transfer in the thin film therefore comprimising the
overall solution.
 \item The wettability coefficient which is in principle can control the
dynamic angle and menisci appearance \cite{pagonabarraga-finger} should not
influence on the overall solution.
 \item The free-energy model examined in the work is one density/different
viscosities model. Classical Bretherton problem is the free-interface problem
where the film thickness is established by itself. The viscosities ratio should
be large as it is indicated in \cite{pagonabarraga-fingers}, however no studies
address this issue. Because the free-interface nature of the Bretherton/Taylor
problem one can have difficulties with imposing the pressure boundary
conditions which do require to impose two separate boundary conditions for two
distribution functions.
 \item A number of simulations are done across the circular shape of channel,
which is quite difficult to address in terms of lattice Boltzmann free energy
framework where one needs to calculate the laplacians near the boundaries.
Although one can apply certain finite difference stencils
\cite{arnold-boundary,hunt-boundary} developed for the curved boundaries
however the analysis of error which can be addressed to the boundary conditions
treatment becomes an issue. 
\end{itemize}
Given all the concerns and challenges the suggested Lattice Boltzmann framework
is the two-dimensional flow driven by gravity or by pressure boundary
conditions. The advantage of the gravity driven bubble is that it's easy to
implement with the periodic boundary conditions. However, the analytical
solution to this problem, to the best authors' knowledge, is not yet known.
Though some authors \cite{sehgal-microchannel} use the gravity driven model to
validate the Bretherton film thickness, the latter should be used with caution
since the velocity of the bubble is different from the fluid velocity and there
is also the additional pressure drop due to front and rear menisci.

>>>>

\section{Introduction}
Taylor/Bretherton \cite{bretherton} flow describes flow of long bubbles in
microchannels. It was found that the front meniscus film width is proportional
to $Ca^{2/3}$ for Capillary numbers less than $0.003$. While there are number
of works which simulate the Bretherton problem with the boundary value method
\cite{ingham-plates,heil-bretherton}, the simulation of Bretherton problem is of
difficulty for diffusion interface methods, where the interface is spread over
the few grid nodes. In this case the question of proper film width resolution
arises. While the discontinuous interface models are good to simulate
non-complicated geometries phenomena, the continuous interface methods are more
flexible to resolve problem evolving coalescence or droplet breakup. This work
concentrates on the resolving parameters range of the binary liquid lattice
Boltzmann method to validate it across the Bretherton problem.

@@

\section{Bretherton bubble flow}
Bretherton in his significant work \cite{bretherton} studied a long bubble
moving in a tube filled with liquid. It was found that the deposition film is
proportional to $Ca^{2/3}$ in the range of small capillary numbers. Later on
it was figured \cite{wong-films,wong-pressure} that the film deposition
proportinal to $Ca^{2/3}$ only near the front meniscus and the film width is
variated through the bubble length for not infinite bubbles. Numerical
simulations \cite{giavedoni-numerical} and experimental studies
\cite{kreutzer-pressure-drop} show as well the deviation of rule $Ca^{2/3}$ for
capillary numbers more than $0.003$ and Reynolds number dependance. To be able
to predict consistently the flow pattern for capillaries in different range of
the parameters researchers validate their codes with the small capillary
numbers Bretherton problem.

There are a number of the numerical methods which
were used for the Taylor/Bretherton flow
simulation. \citet{vanbaten-circular} studied the mass transfer and film
thickness for rising bubbles in a circular capillar.
\citet{kreutzer-pressure-drop} performed numerical simulation by the
finite volume method the circular capillary for the number of different
Reynolds and capillary numbers. \citeauthor{wong-films} in
\cite{wong-films,wong-pressure} studied three-dimensional bubbles in the
polygonal capillaries and showed the derivations of different shapes in the
slug crossections and menisci appearance.
\citet{heil-bretherton,ingham-plates} studied the air finger propagation to
the two-dimensional channel for the range of Reynolds numbers and capillary
numbers. \citet{giavedoni-numerical} performed crossvalidation of the
finite element solution developed by them with the already published results
across the different capillary numbers. The solution is available for circular
and planar cases.

While the applicability of other methods are well established for the
simulation of the Bretherton/Taylor problem, the lattice Boltzmann
applicability and thorough parameters study to the best authors' knowledge is
not yet done. Especially that concerns the diffuse interface multiphase
models, where one needs to resolve the film accordingly. One should acknowledge
the works of
\citet{pagonabarraga-fingers} on meniscus in thin films for the fingering
phenomena. \citet{sehgal-microchannel} performed the lattice Boltzmann
simulations of two-dimensional channel flow for larger capillary numbers where
the authors found the discrepancies with the classical Bretherton theory having
limitations to the low capillary numbers regime \cite{giavedoni-numerical}.

While to resolve interface on a fine level for capillary numbers less than
$0.003$ is computationally extensive even for a two-dimensional case. For
instance for capillary number $0.01$ to properly resolve the interface one
needs around 22 GB memory requirement. Therefore, this work is based on the
comparison with the already established results which were in their case
validated over the Bretherton bubble flow. The work presents recipes for the
initialization and the optimal parameter range for the binary liquid LBM model,
which is described in the next section.

\section{Lattice Botlzmann binary liquid model}
The lattice Boltzmann equation operates on the rectangular grid representing the
physical domain. It utilizes the
probability distribution functions containing the information about the
macroscopic variables, such as density,
momentum. LBE consists from two parts: local collision and propagation operated
from one node to another in the certain
direction specified by the velocity set.  The lattice Boltzmann implementation
in a standard view representing the
collision and propagation steps:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{ collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $F_i$ is the external force population. The binary liquid LBM model is
based on the free-energy functional \cite{swift,landau} and operates with two
sets of populations one to track pressure and velocities fields, another one is
for the phase field. The model to be utilized is two-dimensional nine-velocity
D2Q9 model.

In terms of D2Q9 the equilibrium populations are as \cite{pooley-contact}:
\begin{equation}
\begin{aligned}
&f_i^{eq}=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr), i=1\div8\\
&f_0^{eq}=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}=w_i(\Gamma \mu + \frac{\phi c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi^m
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}), i=1\div8 \\
&g_0^{eq}=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is related to the surface
tension; $A$ is the parameter of the free-energy model; bulk pressure
$p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$. D2Q9 related parameters are
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$ with
the sound speed parameter $c_s^2=\dfrac{1}{3}$. Weights related to the
inclusion of the surface tension coefficient are as follows:
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$. Note that the system restores the macroscopic
equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
&\rho\left(\partial_t+u_{\beta}\partial_{\beta}\right) u_{\alpha}=
-\partial_{\alpha}P_{\alpha \beta} +
\nu\partial_{\beta}\left(\partial_{\alpha}u_{\beta}+\partial_{\beta} u_{\alpha}
+ \frac{1}{3}\partial_{\gamma} u_{\gamma} \delta_{\alpha\beta}\right)\\
&\partial_t \phi + \partial_{\alpha} \phi u_{\alpha}=M \Delta \mu,
\end{aligned}
\label{binary:fluid:system}
\end{equation}
where the viscosity $\nu=c_s^2 (\tau-1/2)$ and mobility parameter
$M=\Gamma(\tau_{\phi}-1/2)$. The system allows the separation of the liquid
phase with $\phi=1$ and so-called gas phase with $\phi=-1$. Note that the
spatial variation of the relaxation parameter depends on the relaxation
parameter of $\tau_{gas}$ and $\tau_{liq}$ linearly as
$\tau=\tau_{gas}+\frac{\phi+1}{2}\tau_{liq}$.

\section{Lattice Botlzmann benchmark}
To properly simulate and to compare simulations results with the already
performed simulations one needs to design a benchmark by addressing the
following challenges for the lattice Boltzmann method:
\begin{itemize}
 \item The diffusive interface nature of the multiphase model should not
influence the film resolution.
 \item The spurious currents arising from the surface tension discretization
can influence the mass transfer in the thin film therefore comprimising the
overall solution.
 \item The wettability coefficient which is in principle can control the
dynamic angle and menisci appearance \cite{pagonabarraga-finger} should not
influence on the overall solution.
 \item The free-energy model examined in the work is one density/different
viscosities model. Classical Bretherton problem is the free-interface problem
where the film thickness is established by itself. The viscosities ratio should
be large as it is indicated in \cite{pagonabarraga-fingers}, however no studies
address this issue. Because the free-interface nature of the Bretherton/Taylor
problem one can have difficulties with imposing the pressure boundary
conditions which do require to impose two separate boundary conditions for two
distribution functions.
 \item A number of simulations are done across the circular shape of channel,
which is quite difficult to address in terms of lattice Boltzmann free energy
framework where one needs to calculate the laplacians near the boundaries.
Although one can apply certain finite difference stencils
\cite{arnold-boundary,hunt-boundary} developed for the curved boundaries
however the analysis of error which can be addressed to the boundary conditions
treatment becomes an issue. 
\end{itemize}
Given all the concerns and challenges the suggested Lattice Boltzmann framework
is the two-dimensional flow driven by gravity or by pressure boundary
conditions. The advantage of the gravity driven bubble is that it's easy to
implement with the periodic boundary conditions. However, the analytical
solution to this problem, to the best authors' knowledge, is not yet known.
Though some authors \cite{sehgal-microchannel} use the gravity driven model to
validate the Bretherton film thickness, the latter should be used with caution
since the velocity of the bubble is different from the fluid velocity and there
is also the additional pressure drop due to front and rear menisci.
